---
title: "EDA&Model"
author: "HongnaSu"
date: "2024-11-21"
output: pdf_document
---

```{r echo=FALSE}
library(dplyr)
library(ggplot2)
library(WDI)
library(rstanarm)
```

#EDA

data clean 
GDP
```{r}

gdp<- WDI(country = "NZ", indicator = "NY.GDP.MKTP.CD", start = 1983, end = 2022)
colnames(gdp) <- c("1","2","3", "year","gdp")
gdp1<-gdp%>% select(-1,-2,-3)
#reverse
gdp2<- gdp1[nrow(gdp1):1, ]
gdp_nz<-data.frame(gdp2)

```
income
```{r}
income<- WDI(country = "NZ", indicator = "NY.GDP.PCAP.CD", start = 1983, end = 2022)
colnames(income) <- c("1","2","3", "4","income")
income1<-income%>% select(-1,-2,-3,-4)
#reverse
income2<-income1[nrow(income1):1, ]
income_nz<-data.frame(income2)
```

imput&output
```{r}
trade<- WDI(country = "NZ", indicator = c("NE.EXP.GNFS.CD", "NE.IMP.GNFS.CD"), start = 1983, end = 2022)
colnames(trade) <- c("1","2","3", "4","input","output")
trade1<-trade%>% select(-1,-2,-3,-4)
trade_nz<-data.frame(trade1)
```

population
```{r}
#Get population data
population <- WDI(country = "NZ", indicator = "SP.POP.TOTL", start = 1983, end = 2022)
colnames(population) <- c("1","2","3", "4","population")
population1<-population%>% select(-1,-2,-3,-4)
#reverse
population2<- population1[nrow(population1):1, ]
population_nz<-data.frame(population2)

```
visitor
```{r}
visitor<- WDI(country = "NZ", indicator = "ST.INT.ARVL", start = 1983, end = 2022)
colnames(visitor) <- c("1","2","3", "4","visitor")
visitor1<-visitor%>% select(-1,-2,-3,-4)
#reverse
visitor2<-visitor1[nrow(visitor1):1, ]
visitor_nz<-data.frame(visitor2)
```
inflation
```{r}
inflation<- WDI(country = "NZ", indicator = "FP.CPI.TOTL.ZG", start = 1983, end = 2022)
colnames(inflation) <- c("1","2","3", "4","inflation")
inflation1<-inflation%>% select(-1,-2,-3,-4)
#reverse
inflation2<-inflation1[nrow(inflation1):1, ]
inflation_nz<-data.frame(inflation2)
```
dataframe
```{r}
#combine all cleaned data
data_nz<-bind_cols(gdp_nz, income_nz,population_nz,trade_nz,visitor_nz,inflation_nz)

summary(data_nz)
```
visualization
```{r echo=FALSE}
  ggplot(data_nz, aes(x =year, y = gdp)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "gdp change from1983-2023", x = "year", y = "gdp")
  
  ggplot(data_nz, aes(x =year, y = income2)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "income change from1983-2023", x = "year", y = "income")
  
  ggplot(data_nz, aes(x =year, y =population2)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "population change from1983-2023", x = "year", y = "population")
  
  ggplot(data_nz, aes(x =year, y = visitor2)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "number of visitor change from1983-2023", x = "year", y = "visitor")
  
  ggplot(data_nz, aes(x =year, y =input)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "input change from1983-2023", x = "year", y = "input")
  
  ggplot(data_nz, aes(x =year, y =output)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "input change from1983-2023", x = "year", y = "output")
  
  ggplot(data_nz, aes(x =year, y =inflation2)) +
  geom_point() +
  geom_smooth(color = "blue")
  theme_minimal() +
  labs(title = "inflation from1983-2023", x = "year", y = "inflation")
```
As we can see from the graphs,GDP,income,population,input,output are linear,which are upwards,and inflation show downwards trend.For visitor,standard error is too big,there are a lot of missing data,especially there is a outlier 2020(because of Covid) makes the curve downwards.These may affect the analyse,we will see in regression part.


Model

This should be a linear problem,so I will consider linear model.

linear model
```{r}
regression1<-lm(gdp~income2+population2+input+output+visitor2+inflation2,data_nz)
summary(regression1)
```
From T test we can see intercept,GDP,income is small enough to indicate strong effect,for but rest of them might not enough,I might choose another model.

Then I use GLM and I remove visitor
```{r echo=FALSE}
regression<-stan_glm(gdp~income2+population2+input+output+inflation2,data_nz,family = gaussian)
summary(regression)
```
Since Rhat = 1.0, it indicates that the model has converged and the model's stability is good.

Prediction
First I get 2023 economic informations
```{r}
data_2023_nz<-data.frame(income2=48527.83,population2=5223100,input=50770000000,output=42210000000,inflation2=5.733163)
```

```{r echo=FALSE}
predicted_GDP_2023<- posterior_predict(regression, newdata = data_2023_nz)
gdp_2023<-mean(predicted_GDP_2023)
gdp_2023
```
Seem its not close,maybe input and output are too much partial in this  model,and in 2023,input and out put decreased a lot,made the model not accuracy.I try to remove input and out put

```{r echo=FALSE}
regression2<-stan_glm(gdp~income2+population2+inflation2,data_nz,family = gaussian)
summary(regression)
```
```{r}
predicted_gdp_2023<- posterior_predict(regression2, newdata = data_2023_nz)
GDP_2023<-mean(predicted_gdp_2023)
GDP_2023
```
This model is better

Prediction of 2024 GDP
```{r}
data_2024_nz<-data.frame(income2=47549.65,population2=5338500,inflation2=5.756)
```

```{r}
predicted_gdp_2024<- posterior_predict(regression2, newdata = data_2024_nz)
GDP_2024<-mean(predicted_gdp_2024)
GDP_2024
```

```{r}

write.csv(data_nz, file = "data_nz.csv", row.names = FALSE)
```


```{r}
library(ggplot2)
library(tidyr) 
df_long <-data_nz%>%
  gather(key = "Variable", value = "Value", gdp,income2, population2,input,output,visitor2,inflation2)
ggplot(df_long, aes(x = year, y = Value, color = Variable)) +
  geom_line(size = 1) + 
  theme_minimal() + 
  labs(title = "Time Series of GDP, Exports, Imports, and Income",
       x = "Date", y = "Value") +
  theme(legend.title = element_blank())
```